Improving Emotion Detection through Real-Time Translation of Text to ML Models Trained in Different Languages

### 2023-08-14 paper/main.tex
Renaming for Submission

\usepackage{array}
\newcolumntype{L}{>{\centering\arraybackslash}m{1.2in}}
\newcolumntype{M}{>{\centering\arraybackslash}m{0.85in}}

@@

\begin{abstract}
This Qualifying Exam\footnote{PhD students of MTSU Computational and Data Science program are required to complete a Qualifying Exam in their first year in the program.} research paper investigates the potential of improving Emotion Detection (ED) through real-time translation of text to multiple Machine Learning (ML) models trained in different languages specifically  English and German. The research aims to address the challenges posed by the lack of comprehensive labeled datasets and language fragmentation in ED analysis research and projects.

By extending an original English language dataset with a translated dataset from German to English the training data (extended dataset) the hope is that better prediction rates can be achieved in ED analysis applications. Furthermore, translating English to German to extend the German dataset and training another ML model in parallel using PySpark and accessing both models in real-time by a simple RESTful API may improve the prediction rates by dual processing a sentence at the same time.

In order to present the results presented both orally to the MTSU's Computational Data \& Science Committee and within this paper labeled datasets both in English and German were collected, translated, and two ML models built that can be accessed via an API to get a prediction on an input (POST or GET) sentence both in German or English.
\end{abstract}
\clearpage

@@

Emotion detection has gained significant interest in recent years as a means to understand human behavior and improve communication effectiveness. By analyzing text or speech, these detection processes can accurately identify and classify emotions expressed by individuals or groups, often times in real-time.

By employing a variety of ED processes, including natural language processing and machine learning, applications can detect emotions with high precision. These ED systems find applications in diverse fields, such as sentiment analysis, customer feedback analysis, voting and news stances, and mental health monitoring, which all aim at contributing to a deeper understanding of human emotions and their impact on various aspects of life.

@@

The concept and number of emotions can vary depending on the theoretical framework or model being considered. One well-known model is the Plutchik's Wheel of Emotions Figure \ref{fig:plutchik-wheel}, which proposes eight (8) primary emotions and their contrasting pairs. In this paper we will be focusing on the basic eight emotions and not on their sub-emotions.

Below are the eight primary emotions according to the Plutchik's Wheel of Emotions \cite{Tromp}, a detailed reference of emotions used in this research can be found in Table \ref{table:dataset_labels} on page~\pageref{table:dataset_labels}.

@@

It's important to note that emotions are complex and multifaceted, and different models may propose alternative categorizations or variations in the number of primary emotions. However for the scope of this research project the most common eight (8) emotions are used as described above.

By aligning the emotional labels across the English \cite{english-dataset-twitter} and German \cite{german-dataset-cheese} datasets being used in this research all eight listed above were used. Details on the linkage of the two datasets is described in Section \ref{sec:data-procurement-process} with an emphasis on the Table \ref{table:dataset_labels}'s cross reference on page~\pageref{table:dataset_labels}.

\subsection{Significance of Emotion Detection (ED) in Machine Learning}
Emotion Detection (ED) analysis in Machine Learning (ML) models has gained significant attention due to its potential applications in various fields, including customer sentiment analysis, stance detection in regard to a specific targets such as news and voting\cite{mascarell-etal-2021-stance}, mental health monitoring\cite{Colonnello}, and human-computer interactions (chatbots\cite{chatbot-cognitive-awareness}) to just name a few.

By accurately identifying and understanding emotions from text data, ML applications can assist in improving user experiences, decision-making processes, and overall human-machine interactions in a positive manner\cite{Colonnello, mascarell-etal-2021-stance}. As an example, the primary goals of chatbots are entertainment, social contact, and novelty interaction, with a strong emphasis on productivity, scale-ability and cost reductions. In chatbots, there has been a immense requirement to simulate human-like characteristics and behavior during human-machine interactions\cite{emotion-detection-literature-review}. Per Kusal et al. \cite{chatbot-cognitive-awareness} these chatbots also improve customer engagement by offering friendliness, comfortness, flexibility, and efficient assistance. Chatbots should provide customers with more engaging responses, directly addressing their issues. The goal of deploying ED applications is for users perceive chatbots as companions rather than simple assistants. Per Adikari et al. the majority of user requests are emotional rather than informative. Chatbots gained the capacity to respond emotionally to customers due to machine learning and sentiment analysis evolution\cite{chatbot-cognitive-awareness}.

@@

Due to the before mentioned wide scope of uses for ED this field of research and the work to advance it provides significance and importance in enabling machines to comprehend and respond appropriately to human interactions. This in turn contributes to the advancement of artificial intelligence, natural language processing and in many cases the utilization of text-to-speech applications most all of us in today's society use and rely on \cite{ai-framework-detection-emotions}.

Lastly, in Chen et al.\cite{research-emotion-recognition-for-online-learning} paper based on emotion detection for online learning, it states that with emotion recognition research, we need to focus on accuracy and real-time performance in order to apply emotional recognition based on physiological signals to solve practical problems; they key part of their claim is that in order for ED to be practical real-time methods needs to be used.

\subsection{ED challenges: Lack of Labeled Datasets and Language Fragmentation}
ED is an immensely growing field, however unlike Sentiment Analysis (SA) the availability of large datasets for training purposes of ML models is much smaller \cite{ACLU-ED-Data, ai-framework-detection-emotions}. The primary reason of the lack of data stems from the emotion detection data requiring primarily supervised learning dataset, which require time and effort to procure, and mostly created and labeled by humans. To compound on the issue of the lack of data; the many datasets that are available are in multiple languages. A significant amount of time took place to find the German labeled

@@

This paper is significant because an entire industry of automated emotion reading (Text, Image, Video, and Sound) technologies is quickly emerging \cite{ACLU-ED-Data, ACLU-THE-DAWN-OF-ROBOT-SURVEILLANCE}. The market for emotion recognition software is forecast to reach at least \$3.8 billion by 2025 \cite{ACLU-ED-Data, ACLU-THE-DAWN-OF-ROBOT-SURVEILLANCE}. And since ED is already being incorporated into products for purposes such as marketing, robotics, driver safety, customer service and a multiple of others, it is fitting to work on research to solve issues with prediction rates and lack of teaching / learning data.

With the before mentioned issues regarding the quantity (lack of) and fragmentation of the ED datasets this research was motivated by finding a way to combine different language datasets into a single set for training processes to improve the predictability of ED. The hope is that by use of translation that the predictability of ED can be improved by use of real-time language conversion and processing.

@@

\begin{itemize}
\item Can by translating English data to German and extending an original German dataset increase the predictability of the ED model?
\item Can by translating German data to English and extending an original English dataset increase the predictability of the ED model?
\item Can by translating in real-time an input to multiple languages improve the predictability based on the combined output of two model's results.
\end{itemize}

\clearpage
\section{Literature Review}
To ensure comprehensive coverage of the latest developments and research in emotion detection and analysis for this research paper, consideration and review was only given to incorporating research papers and online news articles that were published after the year 2015, thereby providing an up-to-date perspective on the subject matter. By implementing the 2015 boundary, our methodology aims to provide a more up-to-date outlook on the subject matter at hand and what areas the ED field in moving in in respect to ML applications and processes.

\subsection{Emotion Detection Importance and Diverse Populations}
Emotion Detection (ED) holds significant importance in various domains, including psychology, social sciences, and human-computer interaction. It enables the analysis of human emotions at scale, providing valuable insights into individual and collective emotional states. In Chowanda et. al. paper\cite{CHOWANDA-2021821} they believe that "Emotions hold a paramount role in the conversation, as it expresses context to the conversation."

Furthermore, emotions can differ across age groups, genders, cultures, and languages. Including data from diverse populations helps in developing inclusive and culturally sensitive ED models. It ensures that the models are not biased towards specific groups and can accurately detect emotions in a wide range of individuals.

\subsection{Lack of Labeled Datasets and Fragmentation caused by Different Languages}

The lack of diverse and large-scale data hinders the development and training of accurate ED models, limiting their effectiveness and general ability across different contexts and populations.

In recent research, efforts have been made to address these limitations by exploring new techniques for emotion detection. For instance, a study by Kapoor\cite{KAPOOR2023120882} et al. proposes analyzing the entire spectrum of the language source to predict emotional changes. This provided some help, but quickly established itself as time consuming and need of expertise in the subject matters. Overcoming the scarcity of labeled datasets and embracing innovative approaches like the before mentioned study by Kapoor\cite{KAPOOR2023120882} et al. and their analysis can contribute to the advancement of ED research. Such advancements hold the potential to enhance the accuracy and applicability of emotion detection models, enabling a deeper understanding of human emotions and more effective responses in various domains.

The paper by Mascarell\cite{mascarell-etal-2021-stance} et al. states that unlike images language operate very differently, therefore the lack of labeled datasets poses a significant challenge in text-based Emotion Detection (ED) analysis as it limits the availability of reliable training data for ML models. This scarcity hampers the model's ability to learn and generalize emotions effectively.

Additionally, Kusal et al. further stipulates \cite{kusal} that the fragmentation caused by different languages further exacerbates the issue, as it reduces the size and diversity of data available for training, resulting in limited cross-lingual generalization and potentially biased models. Overcoming these challenges is crucial for advancing ED research and improving the accuracy and applicability of emotion detection models across various languages and cultures.

@@

\subsection{Data procurement process of ED datasets in English and German}
\label{sec:data-procurement-process}
The data procurement was relatively straight forward and in essence entailed using Google search terms for the Emotion Detection in English and German. Some of the main German data was obtained from the dataset built by ETH's Emotion and Stance Detection for German Text \cite{mascarell-etal-2021-stance} which also hosts a good website with their findings \footnote{https://mtc.ethz.ch/research/natural-language-processing/emotion-stance.html}. The English dataset was downloaded from Kaggle \footnote{https://www.kaggle.com/datasets/pashupatigupta/emotion-detection-from-text} based on Tweets collected in 2021. The English dataset contains over 38,000 tweets and their corresponding emotion labeled. Similarly the German contains over 12,000 sentences and their respective emotions associated.

An important part of the process is to make sure we match the English emotions to those of the German language emotions. Since the tow languages are dissimilar we opted to match exact emotions to each other that can be seen in Table \ref{table:dataset_labels} below.

@@

    \hline
    \multicolumn{2}{|c|}{English} & \multicolumn{2}{|c|}{German} & \multirow{2}{*}{Used} \\
    Name & Count & Name & Count \\
    \hline
    Boredom    &  179 & ---           & --- & NO  \\
    Love       & 3842 & Vertrauen     & 316 & YES \\
    Relief     & 1526 & ---           & --- & NO  \\ 
    Fun        & 1776 & ---           & --- & NO  \\
    Hate       & 1323 & Ekel          &  29 & YES \\
    Neutral    & 8638 & Unklar        & 314 & YES \\
    Anger      &  110 & Ärger         & 226 & YES \\
    Happiness  & 5209 & Freude        & 140 & YES \\
    Surprise   & 2187 & Überraschung  & 369 & YES \\
    Sadness    & 5165 & Traurigkeit   & 184 & YES \\
    Worry      & 8459 & Angst         & 154 & YES \\
    Enthusiasm &  759 & Antizipation  & 774 & YES \\
    Empty      &  827 & ---           & --- & NO  \\
    \hline
\end{tabular}
\caption{Dataset of English \& German and their respective label counts.}
\label{table:dataset_labels}
\end{table}

@@

The procurement of two ED datasets is needed. The first will be in English and the second in German. Each dataset will be translated into the other’s language for extending purposes. As an example Figure \ref{fig:dataset-extension} depicts the original English dataset will be extended by the translated German dataset.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.35\textwidth]{Dataset-Extension}
    \caption{Example of English dataset extension by translating from German.}
    \label{fig:dataset-extension}
\end{figure}

The reason for selecting German was based on the author\footnote{Richard Hoehn} being fluent in English and German and hence being able to easily traverse the two languages while working on the code for dataset translations and formatting.

@@

\subsection{Development of a translation application for converting different languages into English and vice versa}
Translation of the emotion datasets was performed with Google Translator based on the Medium article\cite{Nidhaloff_how_to_translate_text_with_python} in which the author describes option for simple API translations. As an example the following is a brief introduction to Google Translator's API:

\begin{quote}
\begin{minted}[fontsize=\small]{python}
# Using Deep Translator to leveage Google Translate
# Link: https://cloud.google.com/translate/docs/reference/libraries/v2/python
from deep_translator import GoogleTranslator

@@

translated_back = GoogleTranslator(source='auto', target='en').translate(translated)
print(translated_back) # Chocolate milk tastes much better through a straw.
\end{minted}
\end{quote}

Based on the above example the CSV and JSON datasets were individually loaded, parsed, and translated to their respective counter-part languages. Details of the application can be found in the Appendix \ref{appendix:dataset_english} for English and Appendix \ref{appendix:dataset_german} for the German. We decided to only use 1,500 row for each language so since the German (DE\footnote{DE = German}) dataset was not much larger than that. This meant that the English (EN\footnote{EN = Enlgish}) datasets was randomly paired down to 1,500 to match that of the German.

@@

\subsection{Training Multiple ML models with PySpark}

 Using the datasets from Tables \ref{table:pd_de_translated}, \ref{table:pd_en_translated}, and \ref{table:ml_models} that list an approximate $\sim$1,275 and $\sim$2,775 features for training. The algorithms tested in this research project ranged from the traditional machine learning to deep learning techniques, they are:
 \begin{itemize}

@@

Each of these algorithms used and their results are listed in the Results and Analysis Section \ref{sec:results-and-analysis} below. In Chowanda et al. paper\cite{CHOWANDA-2021821} on Text-based Emotion Techniques multiple algorithms were proposed, this research focused on just three out of their six since the others had lower prediction rates based on Chowanda et al. paper\cite{CHOWANDA-2021821} paper.

 \begin{table}[h!]
\centering
\begin{tabular}{ | L | L | L | L | }
    \hline
    \multicolumn{4}{|c|}{ML Models } \\
    \hline
    Model & 
    Data &
    Rows for Training (85\%) & 
    Rows for Testing (15\%)  \\
    \hline
    A &
    English Original &
    1,275 & 
    225 \\
    \hline
    B &
    English Extended By German &
    2,775 & 
    225 same as Model "A" \\
    \hline
    C &
    German Original &
    1,275 & 
    225 \\
    \hline
    D &
    German Extended By English &
    2,775 & 
    225 same as Model "C" \\

    \hline
\end{tabular}
\caption{ML Models and their respective settings}
\label{table:ml_models}
\end{table}

@@

\item \textbf{F1 Score:} Precision represents the proportion of correctly predicted emotions among all instances predicted as a specific emotion with respect to the labels. Recall measures the proportion of correctly predicted emotions among all instances that truly belong to a specific emotion. With these two the F1 Score ($F_{1}=\frac{\text{Precision} \times \text{Recall}}{\text{Precision} + \text{Recall}}$) combines precision and recall into a single metric, providing a balanced evaluation.

@@

While evaluating the models we focused on the Accuracy and the F1 Score to illustrate the prediction results of the original versus extended datasets.

>>>>

\usepackage{array}
\newcolumntype{L}{>{\centering\arraybackslash}m{1.2in}}
\newcolumntype{M}{>{\centering\arraybackslash}m{0.85in}}
\usepackage{tablefootnote}

@@

\begin{abstract}
This Qualifying Exam\footnote{PhD students of MTSU Computational and Data Science program are required to complete a Qualifying Exam in their first year in the program.} research paper investigates the potential of improving Emotion Detection (ED) through real-time translation of text to multiple Machine Learning (ML) models trained in different languages. Our research focused on  English and German datasets. This research aims to address the challenges posed by the lack of comprehensive labeled datasets and language fragmentation in ED analysis research and projects, essence the lack of data to better train ML models is trying to be addressed.

By extending an original English language dataset with a translated dataset from German to English the training data (extended dataset) become larger and the hope is that better prediction rates can be achieved in ED analysis applications due to ML training having more data for learning. Furthermore, translating English to German to extend the German dataset and training another ML model using PySpark and accessing both models in real-time by a simple RESTful API may improve the prediction rates by dual processing a sentence at the same time.

In order to present the results presented both orally to the MTSU's Computational Data \& Science Committee and within this paper labeled datasets both in English and German were collected, translated, and two ML models built that can be accessed via an API to get a prediction on an input (POST or GET) sentence both in German or English.

Sadly the research and results point to the realization that extending the datasets by translation did not improve prediction rates of English or German. Some minor gains can be achieve by accessing both English and German models at the same time via the RESTful API but unfortunately the benefits may not really support the efforts.
\end{abstract}
\clearpage

@@

Emotion detection has gained significant interest in recent years as a means to understand human behavior and improve communication effectiveness. By analyzing text, speech or facial expressions, these detection processes try to accurately identify and classify emotions expressed by individuals or groups, often times in real-time.

By employing a variety of ED processes, including natural language processing (NLP) and machine learning (ML), applications with the right amount of data can detect emotions with high precision. These ED systems find applications in diverse fields, such as sentiment analysis, customer feedback analysis, voting and news stances, and mental health monitoring, which all aim at contributing to a deeper understanding of human emotions and their impact on various aspects of life.

@@

The concept and number of emotions can vary depending on the theoretical framework or model being considered. One well-known model is the Plutchik's Wheel of Emotions Figure \ref{fig:plutchik-wheel}, which proposes eight (8) primary emotions and their contrasting pairs. In this research paper we will be focusing on the basic eight emotions and not on their sub-emotions or contrasting pairs. This allows the ML models being trained and built to be a multi-classification model with eight label options.

Listed below are the eight primary emotions according to the Plutchik's Wheel of Emotions \cite{Tromp} that we will be focusing on both in English and German, a detailed reference of emotions used in this research can be found in Table \ref{table:dataset_labels} on page~\pageref{table:dataset_labels}.

@@

It's important to note that emotions are complex and multifaceted, and different emotion models or frameworks may propose alternative categorizations or variations in the number of primary emotions. However for the scope of this research project the most common eight (8) emotions are used as described above. This allows the ML models being trained and built to be a multi-classification model with eight label options.

By aligning the emotional labels across the English \cite{english-dataset-twitter} and German \cite{german-dataset-cheese} datasets that are being used in this research all eight listed above were used. Details on the linkage of the two datasets and their counts are described in Section \ref{sec:data-procurement-process} with an emphasis on Table \ref{table:dataset_labels}'s cross reference on page~\pageref{table:dataset_labels} of German and English languages.

\subsection{Significance of Emotion Detection in Machine Learning}
Emotion Detection (ED) analysis in Machine Learning (ML) models has gained significant attention due to its potential applications in various fields, including customer sentiment analysis, stance detection in regard to a specific targets such as news and voting\cite{mascarell-etal-2021-stance}, mental health monitoring\cite{Colonnello}, and human-computer interactions (chatbots\cite{chatbot-cognitive-awareness}) to just name a few.

By accurately identifying and understanding emotions from text data, ML applications can assist in improving user experiences, decision-making processes, and overall human-machine interactions in a positive manner\cite{Colonnello, mascarell-etal-2021-stance}.

As an example, the primary goals of chat-bots are entertainment, social contact, and novelty interaction, with a strong emphasis on productivity, scale-ability and cost reductions. In chat-bots, there has been a immense requirement to simulate human-like characteristics and behavior during human-machine interactions\cite{emotion-detection-literature-review}. Per Kusal et al. \cite{chatbot-cognitive-awareness} these chat-bots also improve customer engagement by offering friendliness, comforters, flexibility, and efficient assistance. Chat-bots should provide customers with more engaging responses, directly addressing their issues or questions. The goal of deploying ED applications is for users perceive chat-bots as companions rather than simple assistants, which in turn helps the overall process of getting to a positive result. Per Adikari et al. the majority of user requests are emotional rather than informative; therefore the need that chat-bots gain the capacity to respond emotionally to customers due to machine learning and sentiment analysis evolution\cite{chatbot-cognitive-awareness} is very important and yields for many applications that employee ED a way to improve interactions with machines.

@@

Due to the before mentioned wide scope of uses for ED this field of research and the work to advance it provides significance and importance in enabling machines to comprehend and respond appropriately to human interactions. This in turn contributes to the advancement of artificial intelligence, natural language processing (NLP) and in many cases the utilization of text-to-speech applications most all of us in today's society use and rely on \cite{ai-framework-detection-emotions} to go about our daily lives.

Lastly, in Chen et al.\cite{research-emotion-recognition-for-online-learning} paper based on emotion detection for online learning, it states interestingly that with emotion recognition research, we need to focus on accuracy and real-time performance in order to apply ED based on physiological signals to solve practical problems; the key part of their claim is that in order for ED to be practical, real-time methods needs to be used and deployed for maximum results.

\subsection{Challenges: Lack of Labeled Datasets and Language Fragmentation}
Emotion Detection is an immensely growing field, however unlike Sentiment Analysis (SA) the availability of large datasets for training purposes of ML models is much smaller \cite{ACLU-ED-Data, ai-framework-detection-emotions}. The primary reason of the lack of data stems from the face that emotion detection data requires primarily supervised learning data. In in turn requires time and effort to construct, catalog, and procure,; this all is mostly created and labeled by humans. To compound on the issue of the lack of labeled data; the many datasets that are available are in many cases in multiple languages - not all are in English since emotions that are linked to text are contextual in nature. for this research a a significant amount of time took place to find the German labeled data, which eventually was found online by google searching.

@@

This paper is significant because an entire industry of automated emotion reading (Text, Image, Video, and Sound) technologies is quickly emerging \cite{ACLU-ED-Data, ACLU-THE-DAWN-OF-ROBOT-SURVEILLANCE}. The market for emotion recognition and detection software is forecast to reach at least \$3.8 billion by 2025 \cite{ACLU-ED-Data, ACLU-THE-DAWN-OF-ROBOT-SURVEILLANCE}. And since ED is already being incorporated into products for purposes such as marketing, robotics, driver safety, customer service, news and stance analysis, and a multiple of others, it is fitting to work on research to solve and / or improve issues in regards to prediction rates and lack of teaching / learning data.

With the before mentioned issues regarding the quantity (lack of) and fragmentation of the ED datasets this research was motivated by finding a way to combine different language datasets into a single set for training processes to improve the predictability of ED. The aim of the research is that by use of translation that the predictability of ED can be improved by extending an English dataset with translations from German and also process look-ups in real-time language conversion and processing.

@@

\begin{itemize}
\item Can by translating English data to German and extending an original German dataset increase the predictability of German ED model?
\item Can by translating German data to English and extending an original English dataset increase the predictability of English ED model?
\item Can by translating in real-time an input to multiple languages improve the predictability based on the combined output of two models that were trained in English and German to better the prediction results.
\end{itemize}

With these three (3) objectives our approach will be to procure, translate, train, and evaluate benefits of extending datasets by translation to improve emotion detection.

\clearpage
\section{Literature Review}
To ensure comprehensive coverage of the latest developments and research in emotion detection and analysis for this research paper, consideration and review was only given to incorporating research papers and online news articles that were published after the year 2015, thereby providing an up-to-date perspective on the subject matter at hand.

We believe by implementing the 2015 boundary, our methodology and results aim to provide a more up-to-date outlook on the subject matter at hand and what areas the emotion detection field is moving in with respect to machine learning research, applications and processes.

\subsection{Emotion Detection Importance and Diverse Populations}
Emotion Detection (ED) holds significant importance in various domains, including psychology, social sciences, and human-computer interaction. It enables the analysis of human emotions at scale, providing valuable insights into individual and collective emotional states both in real-time but also as a type barometer for measuring sentiments from the past versus the current time. In Chowanda et al. paper\cite{CHOWANDA-2021821} they believe that "Emotions hold a paramount role in the conversation, as it expresses context to the conversation.", this means that emotions are a part of a conversation and with that are needed to ensure valid analysis of a conversation.

Furthermore Kapoor et al. \cite{KAPOOR2023120882} states that, emotions can differ across age groups, genders, cultures, and languages. Including data from diverse populations helps in developing inclusive and culturally sensitive ED models. It ensures that the models are not biased towards specific groups and can accurately detect emotions in a wide range of individuals.

\subsection{Lack of Labeled Datasets and Fragmentation caused by Different Languages}
The lack of diverse and large-scale data hinders the development and training of accurate ED models, limiting their effectiveness and general ability across different contexts and populations. In recent research, efforts have been made to address these limitations by exploring new techniques for emotion detection.

For instance, a study by Kapoor\cite{KAPOOR2023120882} et al. proposes analyzing the entire spectrum of the language source to predict emotional changes. This provided some help, but quickly established itself as time consuming task and need of expertise in the subject matters that could not outweigh the results. Overcoming the scarcity of labeled datasets and embracing innovative approaches like the before mentioned study by Kapoor\cite{KAPOOR2023120882} et al. and their analysis can contribute to the advancement of ED research. Such advancements hold the potential to enhance the accuracy and applicability of emotion detection models, enabling a deeper understanding of human emotions and more effective responses in various domains, although the cost of such might be prohibitive at this time.

In addition, the paper by Mascarell\cite{mascarell-etal-2021-stance} et al. states that unlike images language operate very differently, therefore the lack of labeled datasets poses a significant challenge in text-based Emotion Detection (ED) analysis as it limits the availability of reliable training data for ML models. This scarcity hampers the model's ability to learn and generalize emotions effectively.

And lastly, a paper by Kusal et al. further stipulates \cite{kusal} that the fragmentation caused by different languages further exacerbates the issue, as it reduces the size and diversity of data available for training, resulting in limited cross-lingual generalization and potentially biased models.

Overcoming these challenges, lack of data and language fragmentation, is crucial for advancing emotion detection and analysis research for improving the accuracy, speed, and applicability of ML models across various languages and cultures.

@@

\subsection{Data procurement process of ED datasets in English and German}
\label{sec:data-procurement-process}
The data procurement was relatively straight forward and in essence entailed using Google search terms for the Emotion Detection in English and German text languages. Some of the main German data was obtained from the dataset built by ETH's Emotion and Stance Detection for German Text \cite{mascarell-etal-2021-stance} which also hosts a good website with their findings \footnote{https://mtc.ethz.ch/research/natural-language-processing/emotion-stance.html}. The English dataset was downloaded from Kaggle \footnote{https://www.kaggle.com/datasets/pashupatigupta/emotion-detection-from-text} based on Tweets collected in 2021. The English dataset contains over 38,000 tweets and their corresponding emotion labeled. Unfortunately the German dataset only contains over 2,000 sentences and their respective emotions associated. this imbalance of the two dataset sizes will be corrected in the subsequent sections.

\begin{table}[h!]
\centering
\begin{tabular}{ | c | c | c | }
    \hline
    \multicolumn{3}{|c|}{File Details} \\
    \hline
    Name & Row Count & Type \\
    \hline
    English & 38,000 & CSV  \\
    German  &  2,500 & JSON \\
    \hline
\end{tabular}
\caption{File details of English \& German data.}
\label{table:file-details}
\end{table}

An important part of the process is to make sure we match the English emotions to those of the German language emotions. Since the two languages are dissimilar we opted to match exact emotions to each other that can be seen in Table \ref{table:dataset_labels} below.

@@

    \hline
    \multicolumn{2}{|c|}{English} & \multicolumn{2}{c|}{German} & \multirow{2}{*}{Used} \\
    Name & Count & Name & Count \\
    \hline
    Boredom    &  179 & ---           & --- & NO  \\
    Love       & 3842 & Vertrauen     & 316 & YES \\
    Relief     & 1526 & ---           & --- & NO  \\ 
    Fun        & 1776 & ---           & --- & NO  \\
    Hate       & 1323 & Ekel          &  29 & YES \\
    Neutral    & 8638 & Unklar        & 314 & YES \\
    Anger      &  110 & Ärger         & 226 & YES \\
    Happiness  & 5209 & Freude        & 140 & YES \\
    Surprise   & 2187 & Überraschung  & 369 & YES \\
    Sadness    & 5165 & Traurigkeit   & 184 & YES \\
    Worry      & 8459 & Angst         & 154 & YES \\
    Enthusiasm &  759 & Antizipation  & 774 & YES \\
    Empty      &  827 & ---           & --- & NO  \\
    \hline
\end{tabular}
\caption{Dataset of English \& German and their respective label counts.}
\label{table:dataset_labels}
\end{table}

@@

The processing of two ED datasets to be synchronized to the emotion is important. The first will be in English and the second in German. Each dataset will be translated into the other’s language for extending purposes. And each row will hold the English and German sentence. As an example Figure \ref{fig:dataset-extension} depicts the original English dataset will be extended by the translated German dataset. By this simple process the English dataset got larger for training purposes.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.4\textwidth]{Dataset-Extension}
    \caption{Example of English dataset extension by translating from German.}
    \label{fig:dataset-extension}
\end{figure}

The reason for selecting German in this research project was based on the author\footnote{Richard Hoehn} being fluent in English and German and hence being able to easily traverse the two languages while working on the code for dataset translations and formatting. It was much easier to create the emotion cross reference Table \ref{table:dataset_labels} and also check the translation results due to the dual language benefit.

@@

\subsection{Development of a translation application for converting different languages into English and vice versa}
Translation of the emotion datasets was performed with Google Translator based on the Medium article\cite{Nidhaloff_how_to_translate_text_with_python} in which the author describes options for simple API translations. Google Cloud Translation was use mainly due to it's free nature and ease of use for this project. There are however many other text based API translation services that could have been used and might yield in some circumstance even better results.

As an example the following is a brief introduction to Google Translator's API with the code snippet below depicting the ease of use to translate text via an API.

\begin{quote}
\begin{minted}[samepage, fontsize=\small]{python}
#
# Using Deep Translator to leveage Google Translate
# Link: https://cloud.google.com/translate/docs/reference/libraries/v2/python
from deep_translator import GoogleTranslator

@@

translated_back = GoogleTranslator(source='auto', target='en').translate(translated)
print(translated_back) # Chocolate milk tastes much better through a straw.

\end{minted}
\end{quote}

Based on the above translation example the CSV (English) and JSON (German) datasets were individually loaded, parsed, and translated to their respective counter-part languages. Details of the application can be found in the Appendix \ref{appendix:dataset_english} for English and Appendix \ref{appendix:dataset_german} for the German processing. We decided to only use 1,500 rows from each language. This decision was made since the German (DE\footnote{DE = German}) dataset was not much larger than that once filtering and processing was completed. This meant that the English (EN\footnote{EN = English}) dataset was randomly paired down to 1,500 to match that of the German.

@@

\subsection{Training Multiple ML models with PySpark}
After the completion of translating and processing the Englsih and German datasets we come to training our models. Using the datasets from Tables \ref{table:pd_de_translated}, \ref{table:pd_en_translated}, and \ref{table:ml_models} that list an approximate $\sim$1,275 and $\sim$2,775 features for training.

It is important to note that in order to accurately test and compare the original and the extended datasets to each other we made sure that the test data was the same. to do this we randomly extracted 15\% of the original datasets (English and German) before extending the original with the translated datasets.

\begin{quote}
\begin{minted}[samepage, fontsize=\small]{python}
#
# Split the English and German Dataframes for Training and Testing
# We are using a 15/85 Split
df_en_train, df_en_test = df_en.randomSplit([0.85, 0.15], seed=2023)
df_de_train, df_de_test = df_de.randomSplit([0.85, 0.15], seed=2023)

# Create the Extended Dataframe with Translated Data
# It is important to add the translated to the training data and not all of it
# Using *df_en_train.columns ensure that the columns are matched with the union
df_en_train_extended = df_en_train.union(df_de.select(*df_en_train.columns))
df_de_train_extended = df_de_train.union(df_en.select(*df_de_train.columns))

\end{minted}
\end{quote}

In Chowanda et al. paper\cite{CHOWANDA-2021821} on Text-based Emotion Techniques multiple algorithms were proposed, our research focused on just three (3) out of the six from the paper since the others had lower prediction rates based on the paper\'s conclusions. The algorithms tested in this research project ranged from the traditional machine learning to deep learning techniques, they are:
 \begin{itemize}

@@

Each of these algorithms used and their results are listed in the Results and Analysis Section \ref{sec:results-and-analysis} below.

 \begin{table}[h!]
\centering
\begin{tabular}{ | L | L | L | L | }
    \hline
    \multicolumn{4}{|c|}{ML Models } \\
    \hline
    Model & 
    Data &
    Rows for Training (85\%) & 
    Rows for Testing (15\%)  \\
    \hline
    A &
    English Original &
    1,275 & 
    225 \\
    \hline
    B &
    English Extended By German &
    2,775 & 
    225 same as Model "A" \\
    \hline
    C &
    German Original &
    1,275 & 
    225 \\
    \hline
    D &
    German Extended By English &
    2,775 & 
    225 same as Model "C" \\
    \hline
\end{tabular}
\caption{ML Models and their respective settings}
\label{table:ml_models}
\end{table}

The results of the testing resulted in some interesting parts that can be viewed in Table \ref{table:ml-prediction-result-versus-test-data}. We grouped the label by count from the test data (both English and German) and compared the predicted result counts to those. It became very clear that there was a strong bias on the GLM (Linear Regression) model that essentially proposed that all sentences evalued to the same output both in English and German.

\begin{table}[h!]
\centering
\begin{tabular}{ | M | M | M | M | M | M | }
    \hline
    \multicolumn{6}{|c|}{ML Models Test versus Test Result Counts} \\
    \multicolumn{6}{|c|}{} \\
    \multicolumn{6}{|c|}{English} \\
    \hline
    ID & 
    Label &
    Test Data &
    NB Res & 
    GLM Res  &
    RFC Res \\
    \hline
    0 & boredom     &  0 & 25 \tablefootnote{In the NB a result that cannot be matched is a zero label} &   0 &   0 \\
    1 & love        & 26 &  8 &   3 &  11 \\
    2 & Relief      &  0 & 33 &   0 &   0 \\
    3 & Fun         &  0 &  5 &   0 &   0 \\
    4 & hate        &  9 & 31 &   0 &   0 \\
    5 & neutral     & 50 & 21 & 152 & 101 \\ 
    6 & anger       &  0 & 36 &   0 &   0 \\
    7 & happiness   & 39 & 44 &   5 &  17 \\
    8 & surprise    & 10 &  9 &   0 &   1 \\
    9 & sadness     & 29 &  0 &   2 &  19 \\
    10 & worry      & 46 &  0 &  50 &  63 \\
    11 & enthusiasm &  3 &  0 &   0 &   0 \\
    \hline
    \multicolumn{6}{|c|}{} \\
    \multicolumn{6}{|c|}{German} \\
    \hline
    ID & 
    Label &
    Test Data &
    NB Res & 
    GLM Res  &
    RFC Res \\
    \hline
    0 & boredom     &  0 & 34 &   0 &   0 \\
    1 & love        & 25 &  5 &   0 &  13 \\
    2 & Relief      &  0 & 28 &   0 &   0 \\
    3 & Fun         &  0 & 16 &   0 &   0 \\
    4 & hate        &  5 & 16 &   0 &   2 \\
    5 & neutral     & 30 & 24 &   0 &   5 \\ 
    6 & anger       & 23 & 20 &   1 &   7 \\
    7 & happiness   & 10 & 20 &   0 &   5 \\
    8 & surprise    & 24 & 47 &   0 &  11 \\
    9 & sadness     & 12 &  0 &   0 &   2 \\
    10 & worry      &  9 &  0 & 209 &   4 \\
    11 & enthusiasm & 72 &  0 &   0 & 161 \\
    \hline
\end{tabular}
\caption{Prediction Results Count compared to Test Data Label Counts}
\label{table:ml-prediction-result-versus-test-data}
\end{table}

@@

\item \textbf{F1 Score:} Precision represents the proportion of correctly predicted emotions among all instances predicted as a specific emotion with respect to the labels. Recall measures the proportion of correctly predicted emotions among all instances that truly belong to a specific emotion. With these two the F1 Score ($F1_{Score}=\frac{\text{Precision} \times \text{Recall}}{\text{Precision} + \text{Recall}}$) combines precision and recall into a single metric, providing a balanced evaluation.

@@

While evaluating the models we focused on the Accuracy and the F1 Score to illustrate the prediction results of the original versus extended datasets, and also compared the different algorithms to each other. As an added example simply using the accuracy to evaluate a model can be miss-leading and therefore in our case using the F1 Score, that is directly correlated to the Precision (eq: \ref{eq:precision} and Recall (eq: \ref{eq:recall}) is immensely important to understanding the ones models and finding the best of for the task of Emotion Detection using Original and Extended datasets.
